Case-based Arabic Morphological Analysis

### 2010-11-04 tex/sarf.tex
why u r not there?

% File naaclhlt2010.tex
%
% Contact: nasmith@cs.cmu.edu

@@

\newcommand{\CharCodeIn}[1]{`\CodeIn{#1}'}
\newcommand{\CodeIn}[1]{{\small\texttt{#1}}}
\newcommand{\frl}[1]{\fbox{\RL{#1}}}

@@

%\author{ Jad Makhlouta \\
%  \And
%Hamza Harkous \\
%  American University of Beirut \\
%  {\tt \{jem04, hhh20, fz11\}@aub.edu.lb }
%  \And 
%  Fadi Zaraket 
%}

@@

\transfalse
\begin{figure}[tb]
\center{
\resizebox{.9\columnwidth}{!}
{ \input{psfigs/exhadith.pstex_t}}
\caption{Hadith abstraction example.}
\label{f:exhadith}
}
\end{figure}
% the FSMs for the three words
\transtrue

We validate our hypothesis using a case study from the Islamic 
literature. 
A {\em hadith} is a narration related to the prophet Mohammad
through a {\em sanad} or a sequence of narrators. 
The authentication of a hadith highly depends on the credibility
of each of the narrators as reported in separate biography 
books. 
In this paper we consider the problem of automatically segmenting
a hadith book into narrations, then segmenting each narration into
its content or {\em matn} and its sanad.
We also partition the sanad accurately into the 
separate narrators so that we can later look each one of them 
up in the biography books.

@@

We make the following key contributions. 
\begin{enumerate}
\item {\bf Case-based analysis:}  We designed and implemented
a case-based morphological analyzer where a case-based controller
machine controls and guides the analysis. 
\item {\bf Exhaustive analysis:} Since we have a case-based 
controller that can decide on the fly whether an analysis is 
accepted or not, we can afford to keep all valid analyses modulo
those pruned by the case controller. 
We do that using non-deterministic FSAs. 
\item {\bf Recursive affixes:}
We refine the concatenative analysis of 
Buckwalter~\shortcite{Buckwalter:02}  and use recursive affixes. 
This allows Sarf to perform better and 
maintain compositional part of speech tags.
\item {\bf Hadith case study:}
We evaluated Sarf using the hadith case study. We were able to
efficiently extract the sequence of narrators into a complex
data structure with three levels of hierarchy with high accuracy. 
\end{enumerate}

@@

Sarf builds upon the lexicon of Buckwalter\shortcite{Buckwalter:02}.
It differs from Buckwalter in that it defines a shorter list of affixes
and a longer list of affix compatibility rules to allow recursive 
affixes so that we can have prefix-prefix an suffix-suffix 
concatenations.
This allows Sarf to better maintain and propagate 
the part of speech information
tags associated with each prefix and suffix.
This information is highly 
redundant with Buckwalter and that may lead to consistency issues. 
Sarf keeps all the possible analyses alive where each analysis
corresponds to a set of decisions and positions in the text. 
Buckwalter produces a set of solutions for each token 
and the several analysis threads of the text need to be 
generated as a product of the token solutions.

Sarf is similar to Beesley~\shortcite{Beesley:01} in that it uses
finite-state Automata (FSA). 
Beesley reports that the number of the Automata generated by a compiler
from Xerox rules can not be controlled and also reports a difficulty to 
compose the FSAs into a single FSA framework. 
Sarf solves the problem with a non-deterministic composition of three
FSAs that are the prefix, stem, and suffix FSAs. 
With Sarf it is easy to control the FSAs with a custom user case 
controller. Up to our understanding, Beesley  
requires recompiling the Xerox rules.

@@

Like ElixirFM~\cite{Otakar:07} Sarf builds on the lexicon
of the Buckwalter analyzer. 
We also use deterministic parsing with tries in our analysis 
to run the prefix, suffix and stem FSAs. 
We think that the inferential-realizational approach 
of ElixirFM
that is highly compatible with the Arabic linguistic 
description~\cite{Badawi:04}
can benefit from many features unique to the Arabic language.
Sarf leaves implementing that to the case-based controller
since in many cases the abstraction needed may not need the 
text to be grammatically correct.
%We can compare the case-based controller in Sarf to a high level
%component in the ElixirFM domain specific language.

@@

Sarf encodes the prefix, suffix and stem lexicons into 
three linear FSAs and uses a non-deterministic composition
of the three FSAs to compute all valid morphological analyses.
We first present a running example of Sarf. 
The running example explains
how the non-deterministic composition works and 
illustrates how Sarf implements recursive affixes and
handles ``run-on'' words.
Then we discuss how Sarf builds and composes the different FSAs.

@@

\transfalse
\begin{figure*}[tb]
\center{
\resizebox{1.7\columnwidth}{!}
{ \input{psfigs/FSM.pstex_t}}
\caption{Example composition of the prefix, stem and suffix FSAs.}
\label{f:example}
}
\end{figure*}
% the FSMs for the three words
\transtrue

The compositional FSA diagram in Figure~\ref{f:example}
illustrates the operation of Sarf when parsing the
string \RL{al-lA`bwn}\RL{wsyl`b-h-aa}.
\footnote{ 
\RL{n}%
\RL{w}
\RL{b}
\RL{`}
\RL{a} 
\RL{l}
\RL{l}
\RL{a} %
\RL{a}
\RL{h}
\RL{b}
\RL{`}
\RL{l}
\RL{y}
\RL{s}
\RL{w}
in separate form to ease following the example
}
The square and circle legends are states of the FSA
and the directed edges are transitions driven by an input
letter that matches the label of the edge. 
The square denotes an accept state, 
the circle denotes a regular state, and the reject states
are omitted for clarity. 
Figure~\ref{f:example}(a) represents the prefix FSA and 
Figures~\ref{f:example}(b) and~\ref{f:example}(c) represent the 
stem and suffix FSAs respectively. 
The symbol $\epsilon$ represents an empty string and is 
the source of non-determinism in this example.

We start at the root square state in the prefix FSA
which is an accept state. 
The edge labeled $\epsilon$ connects the prefix 
FSA to the stem FSA and transitions from any accept state
in the prefix FSA to the root state in the stem FSA.
The edge that connects the stem FSA to the suffix FSA
follows the same behavior.

When there are two valid transitions such as \RL{w} 
and $\epsilon$ in the start case, the non-deterministic 
FSA $\Phi$ spawns an exact copy of itself $\Psi$. FSA $\Phi$
makes one transition, and FSA $\Psi$ makes the other. 
Each of the FSAs now represent a valid analysis so far. 
When the FSA is at a state where the next input letter
leads to a reject state, the FSA dies.
The reject states are omitted from Figure~\ref{f:example}
and are denoted by the absence of the corresponding 
letters. 
In our example, if we assume there were no stems that start 
with the letter \RL{w}, $\Psi$ will die then. 
In reality probably $\Psi$ will die when the input string
is at \RL{wsyl`}.

Note that the prefix and affix FSAs allow recursive 
prefixes. For example, \RL{w} will result in an accept state
that transitions to the stem FSA. 
When \RL{s} follows, we move to another accept state in the 
prefix FSA corresponding to the suffix \RL{ws}. 
The same applies to the suffix FSA.

Lets consider FSA $\Phi$ after it consumed \RL{wsy} 
and transitioned into the stem FSA root node. 
Now $\Phi$ will transition with \RL{l`b} to reach an accept 
state. 
Before moving with the letter \RL{b} to the accept state,
$\Phi$ needs to make sure that the stem \RL{l`b} is compatible
with the prefix \RL{wsy}. 
It does that by associating a category value for each accept state. 
Sarf considers the value of the category as part of the state.
Thus each state shown in Figure~\ref{f:example} actually represents
more than one state. 
If the category of \RL{l`b} is compatible with the category of
\RL{wsy} then \RL{b} will move $\Phi$ to an accept state. 
Otherwise, it will move it to a regular or a reject state.

Since $\Phi$ is now in an accept state in the stem FSA, it will
spawn $\Sigma$ that transitions into the root of the 
suffix FSA. When \RL{h} is consumed, $\Phi$ will die since 
there is no edge from the current state labeled with \RL{h}. 
Now $\Sigma$ represents a valid analysis and can proceed.

Note that $\Sigma$ reports a full word at any accept state
by spawning a new Automata using the $\epsilon$ transition
to the prefix FSA.
In Sarf, this only happens with white space, delimiters, 
and non-connecting letters. 
We left that out in Figure~\ref{f:example} for clarity.

Finally, the case-based controller can interfere at any point in the 
analysis to make a decision like ignoring an Automata that 
is not interesting. 
It may also decide to correct the analysis of one Automata
based on the decision of the others.

@@

Many affixes are composed of other affixes. For example,
the prefix \RL{wsy} is composed of three other prefixes
namely \RL{w}, \RL{s}, and \RL{y}.
The POS tag for \RL{wsy} is a concatenation
of the POS tags of each of the three morphemes. 
Keeping these morphemes separate in a list
is redundant and may produce consistency issues. 
We add two lists of compatibility rules $L_{pp}$ and
$L_{xx}$ for prefix-prefix and suffix-suffix compatibility
respectively.
This allows us to keep a shorter list of affixes. 
We need one more additional modification to accommodate 
a category for the morphemes accepted by the $L_{pp}$
and the $L_{xx}$ rules. For that we introduce the concept
of a {\em resulting category}. 
The resulting category in all practical cases
is that of the second morpheme in the pair.

\subsection{Affix linear FSA}
\label{sec:affixFSA}

We analyzed the prefixes and suffixes of 
Buckwalter~\shortcite{Buckwalter:02}
and extracted our refined lists of recursive affixes.
We encoded the lists into an FSA similar to those
in Figure~\ref{f:example}(a) and Figure~\ref{f:example}(c).
The two FSAs are directed acyclic graphs. 
The absence of cycles in these FSAs reduces the 
traversal of the affix FSAs to a linear traversal.
This is computationally superior to the 
approach of Buckwalter where the analyzer considers
all possible substrings of the word in question
and looks it up in the affix tables.

The prefix and suffix FSAs encode the $L_{pp}$ and
$L_{xx}$ compatibility lists.
An accept state in the FSA corresponds to the last letter 
of a prefix in the prefix and suffix lists.
The accept states hold category state information as well
as POS and other tags.

\subsection{Stem linear FSA}
\label{sec:stemFSA}

@@

The stems share lots of substrings. We encoded them in
an efficient double array trie structure~\cite{Aoe:89}. 
The structure is also a linear FSA where the accept
states are the terminal nodes corresponding to the last 
letters in stems. 
This approach is superior to Buckwalter since it consists of
a linear walk in the trie while with Buckwalter we need
a number of hash lookups in the order of all possible partitions
of the string.

\subsection{Non-deterministic composition of FSAs}
\label{sec:ndfsa}

The diagram in Figure~\ref{f:composition} shows an 
abstraction of the non-deterministic composition 
of the prefix, stem and suffix FSAs. 
The circle represents regular states, the square
represents accept states and the triangle represents
reject states. 
Note that the concrete prefix, stem and suffix FSAs
are all linear and do not have cycles as discussed 
is Sections~\ref{sec:stemFSA} and~\ref{sec:affixFSA}.
The cycles are introduced in the abstract machine
for brevity and illustrative purposes.

\begin{figure*}[tb]
\center{
\resizebox{1.8\columnwidth}{!}
{ \input{psfigs/abstract_machine.pstex_t}}
\caption{An abstraction of the non-deterministic composition of 
    the prefix, stem and suffix FSAs.}
\label{f:composition}
}
\end{figure*}

The algorithm \CodeIn{NDSarf} below controls Sarf. 
It takes as input a text string $L$ and a case-based controller
$A$. 
It starts with one FSA $\Phi_1$ similar to the one in 
Figure~\ref{f:composition} and inserts it into a collection
of FSAs $M$. 
The algorithm reads a letter $c$ at a time from $L$
and iterates over all machines in $M$. 
For each $\Phi$ in $M$,
if $\Phi$ is in an accept state and it is in the suffix
phase, then the algorithm checks whether it should report
a full word. 
This happens when $c$ is a white space or a delimiter, 
or when the last letter leading to the current state
was a non-connecting letter.

In all cases, if $\Phi$ is in an accept state, 
it spawns another FSA $\Psi$ and adds it to $M$. 
If $c$ leads to a reject state, then $\Phi$ dies 
and we remove it from $M$. 
Otherwise, $\Phi$ transitions using the $c$ edge.
After iterating over all FSAs in $M$, we invoke the 
controller $A$
and grant it full access and control over
all FSAs in $M$.

@@

\begin{table}[tb]
\centering
\begin{tabular} {p{6cm}}
\begin{Verbatim}[fontsize=\relsize{-1},
frame=topline,framesep=4mm,label=\fbox{NDSarf algorithm},
commandchars=\\\{\}, codes={\catcode`$=3\catcode`_=8}]
NDSarf(string $L$, Controller $A$) 
  FSAVec $M$; -- collection of FSA Automata
  FSA $\Phi_1$;
  $M$.insert($\Phi_1$);
  foreach $c$ in $L$ \{
    foreach $\Phi$ in $M$ \{
      if ( $\Phi$.state.isAccept() ) \{
        if ($\Phi$.isSuffix())
          if ($c$.isWhite()) or 
            $A$.report();
          if ($\Phi$.lastChar().isNonConnecting())
            $A$.report();
        $\Psi$ = $\Phi$.clone();
        $M$.insert($\Psi$); \}

@@

A hadith in Islamic literature is a narration from the prophet Mohammad
related by multiple narrators.
Establishing the authenticity of a hadith is an important task
in Islamic studies. 
%Several researchers~\cite{Hadithopaedia:08} attempted to automate 
%the analysis of the hadith literature. 
In this paper we use Sarf to successfully automate the
analysis of three books of hadith selected 
arbitrarily~\cite{IbnHanbal,AlKulayni,AlTousi}\footnote{We obtained
  the digitized books from online sources such as 
  \href{http://www.yasoob.com/}{http://www.yasoob.com/} and 
  \href{http://www.al-eman.com/}{http://www.al-eman.com/}. }
%  Our understanding is that those books are historical documents with 
 % public IP.}.

@@

We illustrate the hadith case-based controller FSA 
in Figure~\ref{f:hadith}. 
The transitions in the hadith FSA are excited
by inputs from the Sarf morphological analyzer. 
A NAME label means that Sarf encountered a proper name.
An NRC label means that Sarf met a narrator connector such as
\RL{`an} "on behalf of", \RL{.hada_t} "narrated", \RL{qaal} "said", 
\RL{'a_hbar} "told" or one of their derivations. 
The IBN label corresponds to the word \RL{ibn} "the son of" and is a name connector.
The NISBA label corresponds to an adjective that points to a person such 
as \RL{alma.sriy} "the Egyptian".

@@

\begin{figure}[tb!]
\center{
\resizebox{.9\columnwidth}{!}
{ \input{psfigs/hadith.pstex_t}}
\caption{FSA controller of the hadith case study.}
\label{f:hadith}
}
\end{figure}

@@

The FSA reports a valid chain of narrators when a sequence of names
connected by narrator connectors appears. 
It marks the beginning of the hadith with the beginning of the current sequence,
and marks the end of the hadith with the beginning of the next sequence. 
It also marks the chain of narrators as the sequence itself.

@@

\section{Future Work}
\label{sec:future}

>>>>

% File naaclhlt2010.tex
% Contact: nasmith@cs.cmu.edu

@@

\newcommand{\CharCodeIn}[1]{`\CodeIn{#1}'}
\newcommand{\CodeIn}[1]{{\small\texttt{#1}}}
\newcommand{\frl}[1]{\fbox{\RL{#1}}} 
\newcommand{\noArRL}[1]{\arabfalse\RL{#1}\arabtrue} 
\newcommand{\noTrRL}[1]{\transfalse\RL{#1}\transtrue}

@@

\author{ Jad Makhlouta \\
  \And
Hamza Harkous \\
  American University of Beirut \\
  {\tt \{jem04, hhh20, fz11\}@aub.edu.lb }
  \And 
  Fadi Zaraket 
}

@@

\transfalse
\begin{figure}[tb]
\center{
\resizebox{1.2\columnwidth}{!}
{ \input{psfigs/exhadith.pstex_t}}
\caption{Hadith abstraction example.}
\label{f:exhadith}
}
\end{figure}
% the FSMs for the three words
\transtrue

We validate our hypothesis using a case study from the Islamic 
literature. 
A \RL{.hady_t} is a narration related to the prophet Mohammad
through a \RL{sanad} or a sequence of narrators. 
Figure~\ref{f:exhadith} shows an example \noArRL{.hady_t} in Arabic with its 
transliteration and translation. 
The authentication of a \noArRL{.hady_t} highly depends on the credibility
of each of the narrators as reported in separate biography 
books. 
In this paper we consider the problem of automatically segmenting
a \noArRL{.hady_t} book into narrations, then segmenting each narration into
its content or \RL{matn} and its \noArRL{sanad}.
We also partition the sanad accurately into the 
separate narrators so that we can later look each one of them 
up in the biography books. 
For example, the boxes in Figure~\ref{f:exhadith} are proper names and once
connected toghether they form complex names of narrators. For example
narrator $n_1$ has the first name \noTrRL{qtybT} and his father name is
\noTrRL{s`yd} as the word in between \noTrRL{bn} (son of) indicates.

@@

We make the following key contributions. 
\begin{enumerate}
\item {\bf Case-based analysis:}  We designed and implemented
a case-based morphological analyzer where a case-based controller
machine controls and guides the analysis. 
%\item {\bf Exhaustive analysis:} Since we have a case-based 
%controller that can decide on the fly whether an analysis is 
%accepted or not, we can afford to keep all valid analyses modulo
%those pruned by the case controller. 
%We do that using non-deterministic FSAs. 
\item {\bf Recursive affixes:}
We refine the concatenative analysis of 
Buckwalter~\shortcite{Buckwalter:02}  and use define the affixes
recursively. 
This allows Sarf to perform better and 
maintain compositional part of speech tags.
\item {\bf Hadith case study:}
We evaluated Sarf using the hadith case study. We were able to
efficiently extract the sequence of narrators into a complex
data structure with three levels of hierarchy with high accuracy. 
\end{enumerate}

@@

Sarf builds upon the lexicon of Buckwalter\shortcite{Buckwalter:02}.
It differs from Buckwalter in that it defines a shorter list of affixes
and a longer list of affix compatibility rules to allow recursive 
affixes so that we can have prefix-prefix an suffix-suffix 
concatenations.
This allows Sarf to better maintain and propagate 
the part of speech information
tags associated with each prefix and suffix.
This information is highly 
redundant with Buckwalter.% and that may lead to consistency issues. 
Sarf keeps all the possible analyses alive where each analysis
corresponds to a set of decisions and positions in the text. 
Buckwalter produces a set of solutions for each token 
and the several analysis threads of the text need to be 
generated as a product of the token solutions.

Sarf is similar to Beesley~\shortcite{Beesley:01} in that it uses
finite-state Automata (FSA). 
Beesley reports that the number of the Automata generated by a compiler
from Xerox rules can not be controlled and also reports a difficulty to 
compose the FSAs into a single FSA framework. 
Sarf solves the problem with an FSA that is the non-deterministic 
composition of a controller and three
transducers that are the prefix, stem, and suffix tansducers. 
With Sarf it is easy to control the FSA with a custom user case 
controller. Up to our understanding, Beesley  
requires recompiling the Xerox rules.

@@

Like ElixirFM~\cite{Otakar:07} Sarf builds on the lexicon
of the Buckwalter analyzer. 
We also use deterministic parsing with tries in our analysis 
to run the prefix, suffix and stem transducers. 
We think that the inferential-realizational approach 
of ElixirFM
that is highly compatible with the Arabic linguistic 
description~\cite{Badawi:04}
can benefit from many features unique to the Arabic language.
Sarf leaves implementing that to the case-based controller
since in many cases the abstraction needed may not need the 
text to be grammatically correct.
%We can compare the case-based controller in Sarf to a high level
%component in the ElixirFM domain specific language.

@@

Sarf encodes the prefix, suffix and stem lexicons into 
three linear transducers and uses a non-deterministic composition
of the three to compute all valid morphological analyses.
We first present a running example of Sarf. 
The running example explains
how the non-deterministic composition works and 
illustrates how Sarf implements recursive affixes and
handles ``run-on'' words.
%Then we discuss how Sarf builds and composes the different FSAs.

@@

\transfalse
\begin{figure*}[tb]
\center{
\resizebox{1.7\columnwidth}{!}
{ \input{psfigs/FSM.pstex_t}}
\caption{Example composition of the prefix, stem and suffix transducers.}
\label{f:example}
}
\end{figure*}
% the FSMs for the three words
\transtrue

The diagram in Figure~\ref{f:example}
illustrates the operation of Sarf when parsing the
string \RL{al-lA`bwn}\RL{wsyl`b-h-aa}.
\footnote{ 
\RL{n}%
\RL{w}
\RL{b}
\RL{`}
\RL{a} 
\RL{l}
\RL{l}
\RL{a} %
\RL{a}
\RL{h}
\RL{b}
\RL{`}
\RL{l}
\RL{y}
\RL{s}
\RL{w}
in separate form to ease following the example
}
The square and circle legends are states 
and the directed edges are transitions driven by an input
letter that matches the label of the edge. 
The square denotes an accept state, 
the circle denotes a regular state, and the reject states
are omitted for clarity. 
Figure~\ref{f:example}(a) represents the prefix transducer and 
Figures~\ref{f:example}(b) and~\ref{f:example}(c) represent the 
stem and suffix transducers respectively. 
The symbol $\epsilon$ represents an empty string and is 
the source of non-determinism in this example.

We start at the root square state in the prefix transducer
which is an accept state. 
The edge labeled $\epsilon$ connects the prefix 
transducer to the stem transducer and transitions from any accept state
in the prefix transducer to the root state in the stem transducer.
The edge that connects the stem transducer to the suffix transducer
follows the same behavior.

When there are two valid transitions such as \RL{w} 
and $\epsilon$ in the start case, the non-deterministic 
composition $\Phi$ spawns an exact copy of itself $\Psi$. 
$\Phi$ makes one transition, and $\Psi$ makes the other. 
Each of $\Phi$ and $\Psi$ now represent a valid analysis so far. 
When $\Phi$ or $\Psi$ is at a state where the next input letter
leads to a reject state, it dies.
The reject states are omitted from Figure~\ref{f:example}
and are denoted by the absence of the corresponding 
letters. 
In our example, if we assume there were no stems that start 
with the letter \RL{w}, $\Psi$ will die then. 
In reality probably $\Psi$ will die when the input string
is at \RL{wsyl`}.

Note that the prefix and affix transducers allow recursive 
prefixes. For example, \RL{w} will result in an accept state
that transitions to the stem transducer. 
When \RL{s} follows, we move to another accept state in the 
prefix transducer corresponding to the suffix \RL{ws}. 
The same applies to the suffix transducer.

Lets consider $\Phi$ after it consumed \RL{wsy} 
and transitioned into the stem transducer root node. 
Now $\Phi$ will transition with \RL{l`b} to reach an accept 
state. 
Before moving with the letter \RL{b} to the accept state,
$\Phi$ needs to make sure that the stem \RL{l`b} is compatible
with the prefix \RL{wsy}. 
It does that by associating a category value for each accept state. 
Sarf considers the value of the category as part of the state.
Thus each state shown in Figure~\ref{f:example} actually represents
more than one state. 
If the category of \RL{l`b} is compatible with the category of
\RL{wsy} then \RL{b} will move $\Phi$ to an accept state. 
Otherwise, it will move it to a regular or a reject state.

Since $\Phi$ is now in an accept state in the stem transducer, it will
spawn $\Sigma$ that transitions into the root of the 
suffix transducer. When \RL{h} is consumed, $\Phi$ will die since 
there is no edge from the current state labeled with \RL{h}. 
Now $\Sigma$ represents a valid analysis and can proceed.

Note that $\Sigma$ reports a full word at any accept state
by spawning a new composition using the $\epsilon$ transition
to the prefix transducer.
In Sarf, this only happens with white space, delimiters, 
and non-connecting letters. 
We left that out in Figure~\ref{f:example} for clarity.

Finally, the case-based controller can interfere at any point in the 
analysis to make a decision like ignoring a composition that 
is not interesting. 
It may also decide to correct the analysis of one Automata
based on the decision of the others. 
The controller uses the output of the transducers as input
to make its decision. 
It does not produce any output and therefore the composition
of the three transducers with the controller is a finite 
state Automaton (FSA).

@@

Many affixes are composed of other affixes. For example,
the prefix \RL{wsy} is composed of three other prefixes
namely \RL{w}, \RL{s}, and \RL{y}.
The part of speech (POS) tag for \RL{wsy} is a concatenation
of the POS tags of each of the three morphemes. 
Keeping these morphemes separate in a list
is redundant.
When users try to add their custom morphemes to these
lists they may produce consistency issues. 
We add two lists of compatibility rules $L_{pp}$ and
$L_{xx}$ for prefix-prefix and suffix-suffix compatibility
respectively.
This allows us to keep a shorter list of affixes. 
We need one more additional modification to accommodate 
a category for the morphemes accepted by the $L_{pp}$
and the $L_{xx}$ rules. For that we introduce the concept
of a {\em resulting category}. 
The resulting category in all practical cases
is that of the second morpheme in the pair.

Formally we define an affix $f$ recursively to be either
an atomic affix $a$ or an atomic affix concatenated
to an affix and we write $f: a | a f$. 
Atomic affixes are standalone affixes
that are not the concatenation of other affixes. 
Sarf only keeps atomic affixes and their categories.

\subsection{Affix linear transducer}
\label{sec:affixFSA}

We analyzed the prefixes and suffixes of 
Buckwalter~\shortcite{Buckwalter:02}
and extracted our refined lists of recursive affixes.
We encoded the lists into a structure similar to those
in Figure~\ref{f:example}(a) and Figure~\ref{f:example}(c).
The two structures are directed acyclic graphs. 
The absence of cycles reduces the 
traversal of the affix transducer to a linear traversal.
This is computationally superior to the 
approach of Buckwalter where the analyzer considers
all possible substrings of the word in question
and looks it up in the affix tables.

The prefix and suffix transducers encode the $L_{pp}$ and
$L_{xx}$ compatibility lists.
An accept state in the transducer corresponds to the last letter 
of a prefix in the prefix and suffix lists.
The accept states produce as output the POS and other tags
associated with the state.

\subsection{Stem linear transducer}
\label{sec:stemFSA}

@@

The stems share lots of substrings. We encoded them in
an efficient double array trie structure~\cite{Aoe:89}. 
The structure is also linear where the accept
states are the terminal nodes corresponding to the last 
letters in stems. 
This approach is superior to Buckwalter since it consists of
a linear walk in the trie while with Buckwalter we need
a number of hash lookups in the order of all possible partitions
of the string.

\subsection{Non-deterministic composition of transducers}
\label{sec:ndfsa}

The diagram in Figure~\ref{f:composition} shows an 
abstraction of the non-deterministic composition 
of the prefix, stem and suffix transducers. 
The circle represents regular states, the square
represents accept states and the triangle represents
reject states. 
Note that the concrete prefix, stem and suffix transducers
are all linear and do not have cycles as discussed 
is Sections~\ref{sec:stemFSA} and~\ref{sec:affixFSA}.
The cycles are introduced in the abstract machine
for brevity and pedagogical purposes.

\begin{figure*}[tb]
\center{
\resizebox{1.8\columnwidth}{!}
{ \input{psfigs/abstract_machine.pstex_t}}
\caption{An abstraction of the non-deterministic composition of 
    the prefix, stem and suffix transducers.}
\label{f:composition}
}
\end{figure*}

The algorithm \CodeIn{NDSarf} below controls Sarf. 
It takes as input a text string $L$ and a case-based controller
$A$. 
It starts with one machine $\Phi_1$ similar to the one in 
Figure~\ref{f:composition} and inserts it into a collection
$M$. 
The algorithm reads a letter $c$ at a time from $L$
and iterates over all machines in $M$. 
For each $\Phi$ in $M$,
if $\Phi$ is in an accept state and it is in the suffix
phase, then the algorithm checks whether it should report
a full word. 
This happens when $c$ is a white space or a delimiter, 
or when the last letter leading to the current state
was a non-connecting letter.

In all cases, if $\Phi$ is in an accept state, 
it spawns another machine $\Psi$ and adds it to $M$. 
If $c$ leads to a reject state, then $\Phi$ dies 
and we remove it from $M$. 
Otherwise, $\Phi$ transitions using the $c$ edge.
After iterating over all machines in $M$, we invoke the 
controller $A$
and grant it full access and control over
all machines in $M$.

@@

\begin{table}[tb]
\centering
\begin{tabular} {p{6cm}}
\begin{Verbatim}[fontsize=\relsize{-1},
frame=topline,framesep=4mm,label=\fbox{NDSarf algorithm},
commandchars=\\\{\}, codes={\catcode`$=3\catcode`_=8}]
NDSarf(string $L$, Controller $A$) 
  MachineVec $M$; -- collection of machines
  Machine $\Phi_1$;
  $M$.insert($\Phi_1$);
  foreach $c$ in $L$ \{
    foreach $\Phi$ in $M$ \{
      if ( $\Phi$.state.isAccept() ) \{
        if ($\Phi$.isSuffix())
          if ($c$.isWhite()) or 
            $A$.report();
          if ($\Phi$.lastChar().isNonConnecting())
            $A$.report();
        $\Psi$ = $\Phi$.clone();
        $M$.insert($\Psi$); \}

@@

A hadith in Islamic literature is a narration from the prophet Mohammad
related by multiple narrators.
Establishing the authenticity of a hadith is an important task
in Islamic studies. 
Based on that Islamic scholars issue rules that affect the life
of people around the world. 
The literature lacks an exhaustive mechanism of authentication
check as. This is currently manual work prone to error and almost
impossible to complete in the life time of a scholar due to the
large literature available. 
%Several researchers~\cite{Hadithopaedia:08} attempted to automate 
%the analysis of the hadith literature. 
In this paper we use Sarf to successfully automate the
analysis of three books of hadith selected 
arbitrarily~\cite{IbnHanbal,AlKulayni,AlTousi}\footnote{We obtained
  the digitized books from online sources such as 
  \href{http://www.yasoob.com/}{http://www.yasoob.com/} and 
  \href{http://www.al-eman.com/}{http://www.al-eman.com/}. }
%  Our understanding is that those books are historical documents with 
 % public IP.}.

@@

We illustrate the hadith case-based controller 
in Figure~\ref{f:hadith}. 
The transitions in the hadith controller are excited
by inputs from the Sarf morphological analyzer. 
A NAME label means that Sarf encountered a proper name.
An NRC label means that Sarf met a narrator connector such as
\RL{`an} "on behalf of", \RL{.hada_t} "narrated", \RL{qaal} "said", 
\RL{'a_hbar} "told" or one of their derivations. 
The IBN label corresponds to the word \RL{ibn} "the son of" and is a name connector.
The NISBA label corresponds to an adjective that points to a person such 
as \RL{alma.sriy} "the Egyptian".

@@

\begin{figure}[tb!]
\center{
\resizebox{.9\columnwidth}{!}
{ \input{psfigs/hadith.pstex_t}}
\caption{The controller of the hadith case study.}
\label{f:hadith}
}
\end{figure}

@@

The controller reports a valid chain of narrators when a sequence of names
connected by narrator connectors appears. 
It marks the beginning of the hadith with the beginning of the current sequence,
and marks the end of the hadith with the beginning of the next sequence. 
It also marks the chain of narrators as the sequence itself.

@@

\subsection{Comparison with Buckwalter and ElixirFM}

We implemented our case study with Buckwalter in two phases. 
In phase one, we process the entire text with Buckwalter and
obtain annotations to locations in text. 
Then in the second phase we feed these annotations to 
our controller as if they were coming from Sarf. 
We measure the time it took to produce the annotations
and the time it took the controller to process the 
annotations whithout the time spent streaming the annotations.
We used the same setup for ElixirFM.

Sarf outperforms both Buckwalter and ElixirFM in runing time. 
It also produces higher accuracy measures. 
We think this is mainly due to the extended
lexicon as well as to the fact that Sarf reduces false positive
names by ignoring prefixes and suffixes that can not form
names even if the stem was a name. 
Doing the same with Buckwalter and ElixirFM requires  
expanding the annotations resulting from one affix to reflect
its multiple composing affixes.
This is easily done with the way we defined affixes recursively.

Buckwalter performed much better than ElixirFM. This is up to 
our knowledge the first comparison of these two tools and 
we think the big difference is due to the more exact 
and grammar based functional morphological analysis ElixirFM 
uses.

\section{Conclusion and Future Work}
\label{sec:future}

In this paper we o
